### 2010-02-24 kivaSections.tex

\documentclass[12pt]{article}
\usepackage{graphicx,amssymb}

@@

\subsection{Descriptive statistics}

In 2003, three different observers flew over the rivers. Assuming that no tracks 
are laid down between individuals' flights on the same day, all three people 
should have seen the same tracks. While there is some error in classifying which 
site the observer saw the actual track in, each observer should at least be 
seeing the same, or nearly the same, number of sites with tracks. However, we 
did not find that to be the case. If we look only at days that all three people 
observed, there appear to be major differences among the proportion of sites in 
which the scientists detected a track (Figure x). The difference in estimated 
na\"ive occupancy among the observers indicates that the scientists did not 
always correctly identify the tracks.

@@

\subsection{False positive rates}
The lack of consistency among the flights is strong evidence that the data 
contain false positives where scientists think they see a track when it does not 
exist. Not accounting for such a process can lead to major biases; as the number 
of flights increases, the occupancy rate estimate will approach one (Royle and 
Link 2006). We introduced a false detection process and used the observer as a 
covariate (by means of indicator variables) for both the probability of 
correctly identifying a track that exists ($p$) and the probability of 
identifying a track that does not exist ($E$). Including false positives 
introduces a duality whereby $\begin{array}{ccc}(p = x, & E = y, & \psi = z)\end
{array}$ is as good a solution as $\begin{array}{ccc}(p = y, & E = x, & \psi = 
1-z)\end{array}$. To address this problem we assumed $p\gg0.5$ and $E\ll0.5$, an 
approach recommended by Royle and Link (2006).

\section{Spatial correlation of the data}
We found that there is significant correlation between the probability of 
observing a track in neighboring sites, but this violates of a central 
assumption of occupancy theory. This issue needed to be addressed. One approach 
was to make the prior distribution of $\theta$, the probability that an otter 
lays down a track, spatially correlated using a conditional autoregressive
(CAR) model.

>>>>

\documentclass[12pt]{article}
\usepackage{graphicx,amssymb,amsmath}

@@

\subsection{Descriptive statistics}
In 2003, three different observers flew over the rivers. Assuming that no tracks 
are laid down between individuals' flights on the same day, all three people 
should have seen the same tracks. While there is some error in classifying which 
site the observer saw the actual track in, each observer should at least be 
seeing the same, or nearly the same, number of sites with tracks. However, we 
did not find that to be the case. If we look only at days that all three people 
observed, there appear to be major differences among the proportion of sites in 
which the scientists detected a track (Figure x). The difference in estimated 
na\"ive occupancy among the observers indicates that the scientists did not 
always correctly identify the tracks.

@@

\subsection{False positive rates}
The lack of consistency among the flights is strong evidence that the data 
contain false positives where scientists think they see a track when it does not 
exist. Not accounting for such a process can lead to major biases; as the number 
of flights increases, the occupancy rate estimate will approach one (Royle and 
Link 2006). We introduced a false detection process and used the observer as a 
covariate (by means of indicator variables) for both the probability of 
correctly identifying a track that exists ($p$) and the probability of 
identifying a track that does not exist ($E$). Including false positives 
introduces a duality whereby 
$\begin{array}{ccc}(p = x, & E = y, & \psi = z)\end{array}$ 
is as good a solution as 
$\begin{array}{ccc}(p = y, & E = x, & \psi = 1-z)\end{array}$. 
To address this problem we assumed $p\gg0.5$ and $E\ll0.5$, an approach 
recommended by Royle and Link (2006).

\section{Spatial correlation of the data}
One approach to dealing with the significant correlation in the probability of 
observing a track in neighboring sites was to use a conditional autogregressive 
(CAR) prior distribution for $\theta$, the probability that an otter lays down a 
track in a site on a given day.

\subsection{Conditional autoregressive (CAR) models}
Let $\theta_s$ be the probability of a track being laid down in site $s$ on a 
given day. Then the Gaussian CAR model can be defined using the following 
conditional distributions:
\begin{equation}\theta_s|\theta_i,\tau_s^2 \sim N(\mu_s+\sum_{i\in N_s}c_{si}
(y_i-\mu_i),\tau_s^2)\end{equation}
where $N_s$ is the set of neighbors of area $s$ (in this case, the two bordering 
sites), $E(\theta_s)=\mu_s$, $\tau_s^2$ is the conditional variance, $c_{si}$ is 
the weight defining how strong the correlation is between sites $s$ and $i$, and 
$c_{ii}=0$ for all sites $i$ (Arab et al., year?). Essentially, this model means 
that if site $s$'s neighbor has a particularly high or low track-laying 
probability, that will increase or decrease the track-laying probability for 
site $s$. This makes it more likely to have strings of tracks that cross over 
multiple sites.

\subsection{Our CAR model}\label{car model}
For our model, both neighbors are weighted equally. Since $\tau_s$ for any site 
$s$ is unknown, we followed convention and used the same value of $\tau$ for all 
sites and placed a prior distribution on it. We chose a $\gamma(0.5,0.0005)$ 
prior distribution for $(\tau^2)^{-1}$, a common choice (GeoBUGS manual). In our 
model we define $\text{logit}(\theta_s)=\alpha_0+\alpha_{1_s}$ where $\alpha_0$ 
has an improper uniform distribution and $\alpha_1$ has a Gaussian CAR 
distribution. We again used a uniform prior distribution on $\psi$ and the same 
beta prior distributions on $p$ and $E$. In this case, we limited the range of 
$p$ to the interval $(0.7,1)$ and the range of $E$ to $(0,0.3)$. We increased 
the burn-in time to 12,000 iterations. With this model, instead of a single 
estimate for $\theta$ there are now $S$ estimates for $\theta$ where $S$ is the 
total number of sites. While ideally we would estimate an overall value for $
\theta$, it seems more important for the purposes of this investigation to 
estimate $\psi$.

\subsection{Simulating spatially correlated data}
Because we modeled spatial correlation in $\theta$, we began by simulating 
spatial correlation in the track-laying process. To do this, we added an 
additional variable to the simulated data, $\alpha$. Under the new simulation 
method, $\theta$ is the probability of an otter laying down a track in site $i$ 
given site $i$ is occupied and site $i-1$ does \textit{not} have a track. If 
site $i$ is occupied and site $i-1$ does have a track, then the probability of 
an otter laying down a track in site $i$ is equal to $\theta+\alpha$. This means 
if there is a track in one site, there is more likely to be a track in the next site. However, when we ran data simulated using this new method on the model 
described in \ref{car model}, the model almost never converged. We tested 
independence of the data using the $\hat{s}$ statistic we developed (section %
\ref{shat}
) and found that the simulated data displayed much more independence than the 
actual data. We tried simulating spatial correlation only in $\psi$ with similar 
results.

To increase the correlation of the tracks we simulated data with spatial 
correlation in both $\theta$ and $\psi$. Under this simulation method we again 
had the variables $\theta$ and $\alpha$, but we also had analogous variables $
\psi$ and $\beta$. In this case, $\psi$ is the probability an otter occupies 
site $i$ given that site $i-1$ is not occupied and $\psi+\beta$ is the 
probability that an otter occupies site $i$ given that site $i-1$ is occupied. 
With this increased spatial correlation we were able to more closely match the $
\hat{s}$ $z$-scores found in the actual data. In addition, the CAR model 
converged more often with data simulated using this method, but it still did not 
converge consistently.

When we ran the data with spatial correlation in the occupancy process on the 
CAR model, we could no longer compare the model estimate for $\psi$ to the 
simulated value of $\psi$ because they had different meanings. To get a sense of 
the overall occupancy rate in the simulated data we counted the number of sites 
that were occupied (known because the data are simulated) and divided that by 
the total number of sites. This proportion is the value we compared to the $\psi
$ estimate from both hierarchical models.
